---
title: "Heart Disease Risk Analysis: Cholesterol Profiles and Lifestyle Factors"
author: "Sheikh Azam Uddin"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    df_print: kable
  word_document:
    toc: true
  pdf_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(dplyr)
library(ggplot2)
library(reshape2)
library(broom)
library(knitr)
```

# 1. Introduction

## 1.1 Problem Statement

Heart disease remains a leading cause of death globally, and its early detection is vital.
Understanding how cholesterol levels — specifically total cholesterol, LDL (low-density
lipoprotein), and HDL (high-density lipoprotein) — along with factors such as blood
pressure, smoking, and physical activity contribute to heart disease risk is crucial for
prevention strategies.

## 1.2 Research Question

To what extent do serum cholesterol profiles (total cholesterol, LDL, and HDL levels)
independently predict heart disease risk, when also considering blood pressure, smoking,
and physical activity?

## 1.3 Statistical Approach

Logistic regression and hypothesis testing (chi-square tests and a two-sample t-test) are
used to identify and quantify associations between cholesterol profiles, lifestyle factors,
and heart disease status. The logistic regression model estimates each predictor's
contribution while holding the remaining covariates constant.

## 1.4 Dataset

The dataset is sourced from Kaggle and comprises 10,000 patient records, covering
cardiovascular biomarkers (blood pressure, cholesterol, triglycerides) alongside
self-reported lifestyle behaviors (smoking, exercise habits) and clinical diagnoses
(hypertension, HDL/LDL classifications).

**Outcome variable:** `Heart.Disease.Status` — Categorical (Yes / No)

**Predictors used in this analysis:**

| Variable | Type | Description |
|---|---|---|
| `Blood.Pressure` | Numerical | Systolic blood pressure |
| `Cholesterol.Level` | Numerical | Total serum cholesterol |
| `Triglyceride.Level` | Numerical | Serum triglyceride level |
| `High.Blood.Pressure` | Categorical | Clinical hypertension (Yes / No) |
| `Exercise.Habits` | Categorical | Physical activity level (High / Medium / Low) |
| `Smoking` | Categorical | Current smoker (Yes / No) |
| `Low.HDL.Cholesterol` | Categorical | Low HDL flag (Yes / No) |
| `High.LDL.Cholesterol` | Categorical | High LDL flag (Yes / No) |

---

# 2. Data Loading and Preparation

## 2.1 Load Data

```{r load-data}
heart_data <- read.csv("heart_disease.csv")
cat("Dataset dimensions:", nrow(heart_data), "rows x", ncol(heart_data), "columns\n")
kable(head(heart_data[, 1:8]), caption = "First 6 rows (first 8 columns shown)")
```

## 2.2 Variable Selection and Outcome Encoding

```{r select-variables}
names(heart_data) <- make.names(names(heart_data))

num_vars <- c("Blood.Pressure", "Cholesterol.Level", "Triglyceride.Level")
cat_vars <- c("High.Blood.Pressure", "Exercise.Habits", "Smoking",
              "Low.HDL.Cholesterol", "High.LDL.Cholesterol")

heart_data <- heart_data %>%
  select(all_of(num_vars), all_of(cat_vars), Heart.Disease.Status) %>%
  filter(Heart.Disease.Status %in% c("Yes", "No")) %>%
  mutate(Heart.Disease.Status = factor(Heart.Disease.Status, levels = c("No", "Yes")))
```

## 2.3 Missing Value Imputation

```{r impute}
get_mode <- function(x) {
  ux <- na.omit(unique(x))
  ux[which.max(tabulate(match(x, ux)))]
}

heart_data[num_vars] <- lapply(heart_data[num_vars], function(x) {
  ifelse(is.na(x), median(x, na.rm = TRUE), x)
})

heart_data[cat_vars] <- lapply(heart_data[cat_vars], function(x) {
  x <- trimws(as.character(x))
  x[x == ""] <- NA
  x[is.na(x)] <- get_mode(x)
  factor(x)
})

missing_check <- data.frame(
  Variable = names(heart_data),
  Missing  = colSums(is.na(heart_data)),
  row.names = NULL
)
kable(missing_check, caption = "Missing value counts after imputation")
```

Numerical variables were imputed using their **median**; categorical variables were
imputed using the **mode** (most frequent category).

---

# 3. Descriptive Statistics

## 3.1 Numerical Variables

```{r numerical-summary}
num_summary <- data.frame(
  Variable = c("Blood Pressure", "Cholesterol Level", "Triglyceride Level"),
  N        = sapply(heart_data[, num_vars], function(x) sum(!is.na(x))),
  Mean     = sapply(heart_data[, num_vars], function(x) round(mean(x),   2)),
  Median   = sapply(heart_data[, num_vars], function(x) round(median(x), 2)),
  SD       = sapply(heart_data[, num_vars], function(x) round(sd(x),     2)),
  Min      = sapply(heart_data[, num_vars], function(x) round(min(x),    2)),
  Max      = sapply(heart_data[, num_vars], function(x) round(max(x),    2)),
  row.names = NULL
)
kable(num_summary, caption = "Summary statistics for numerical predictors")
```

## 3.2 Categorical Variables

```{r categorical-summary}
all_cat <- c(cat_vars, "Heart.Disease.Status")
for (var in all_cat) {
  freq_df <- as.data.frame(table(heart_data[[var]]))
  names(freq_df) <- c("Category", "Count")
  freq_df$Percentage <- paste0(round(prop.table(freq_df$Count) * 100, 1), "%")
  print(kable(freq_df,
              caption = paste("Distribution of:", gsub("\\.", " ", var)),
              row.names = FALSE))
}
```

---

# 4. Data Visualization

## 4.1 Categorical Predictors vs. Heart Disease Status

```{r categorical-plots, fig.height=4}
plot_labels <- c(
  "High.Blood.Pressure"  = "High Blood Pressure",
  "Low.HDL.Cholesterol"  = "Low HDL Cholesterol",
  "High.LDL.Cholesterol" = "High LDL Cholesterol",
  "Smoking"              = "Smoking Status",
  "Exercise.Habits"      = "Exercise Habits"
)

for (var in names(plot_labels)) {
  p <- ggplot(heart_data, aes(x = .data[[var]], fill = Heart.Disease.Status)) +
    geom_bar(position = "dodge") +
    scale_fill_manual(values = c("No" = "#2196F3", "Yes" = "#F44336")) +
    theme_minimal() +
    labs(
      title = paste("Heart Disease by", plot_labels[[var]]),
      x     = plot_labels[[var]],
      y     = "Number of Individuals",
      fill  = "Heart Disease"
    )
  print(p)
}
```

## 4.2 Numerical Predictors vs. Heart Disease Status

```{r boxplots, fig.height=4}
box_labels <- c(
  "Blood.Pressure"     = "Blood Pressure",
  "Cholesterol.Level"  = "Cholesterol Level",
  "Triglyceride.Level" = "Triglyceride Level"
)

for (var in names(box_labels)) {
  p <- ggplot(heart_data,
              aes(x = Heart.Disease.Status, y = .data[[var]],
                  fill = Heart.Disease.Status)) +
    geom_boxplot(alpha = 0.7) +
    scale_fill_manual(values = c("No" = "#2196F3", "Yes" = "#F44336")) +
    theme_minimal() +
    theme(legend.position = "none") +
    labs(
      title = paste(box_labels[[var]], "by Heart Disease Status"),
      x     = "Heart Disease Status",
      y     = box_labels[[var]]
    )
  print(p)
}
```

## 4.3 Correlation Heatmap

```{r correlation-heatmap, fig.height=4, fig.width=5}
numeric_cols <- names(heart_data)[sapply(heart_data, is.numeric)]
cor_matrix   <- cor(heart_data[, numeric_cols], use = "complete.obs")
cor_melted   <- melt(cor_matrix)

ggplot(cor_melted, aes(x = Var1, y = Var2, fill = value)) +
  geom_tile(color = "white") +
  geom_text(aes(label = round(value, 2)), size = 3) +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white",
                       midpoint = 0, limit = c(-1, 1), name = "Correlation") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Correlation Heatmap of Numerical Variables", x = "", y = "")
```

---

# 5. Statistical Analysis

## 5.1 Chi-Square Test: Low HDL Cholesterol vs. Heart Disease

```{r chisq-hdl}
table_hdl <- table(heart_data$Low.HDL.Cholesterol, heart_data$Heart.Disease.Status)
kable(as.data.frame.matrix(table_hdl),
      caption = "Contingency table: Low HDL Cholesterol x Heart Disease Status")

hdl_test <- chisq.test(table_hdl)

kable(data.frame(
  Statistic   = round(hdl_test$statistic, 3),
  df          = hdl_test$parameter,
  `p-value`   = round(hdl_test$p.value, 4),
  check.names = FALSE
), row.names = FALSE, caption = "Chi-Square Test Result: Low HDL Cholesterol")
```

## 5.2 Chi-Square Test: High LDL Cholesterol vs. Heart Disease

```{r chisq-ldl}
table_ldl <- table(heart_data$High.LDL.Cholesterol, heart_data$Heart.Disease.Status)
kable(as.data.frame.matrix(table_ldl),
      caption = "Contingency table: High LDL Cholesterol x Heart Disease Status")

ldl_test <- chisq.test(table_ldl)

kable(data.frame(
  Statistic   = round(ldl_test$statistic, 3),
  df          = ldl_test$parameter,
  `p-value`   = round(ldl_test$p.value, 4),
  check.names = FALSE
), row.names = FALSE, caption = "Chi-Square Test Result: High LDL Cholesterol")
```

## 5.3 Welch Two-Sample t-Test: Cholesterol Level by Heart Disease Status

```{r ttest-cholesterol}
chol_ttest <- t.test(Cholesterol.Level ~ Heart.Disease.Status, data = heart_data)

kable(data.frame(
  `Mean (No)`   = round(chol_ttest$estimate[1], 2),
  `Mean (Yes)`  = round(chol_ttest$estimate[2], 2),
  `t-statistic` = round(chol_ttest$statistic, 3),
  df            = round(chol_ttest$parameter, 1),
  `p-value`     = round(chol_ttest$p.value, 4),
  check.names   = FALSE
), row.names = FALSE, caption = "Welch Two-Sample t-Test: Cholesterol Level by Heart Disease Status")
```

## 5.4 Logistic Regression Model

```{r logistic-regression}
logit_model <- glm(
  Heart.Disease.Status ~ Cholesterol.Level + Low.HDL.Cholesterol +
    High.LDL.Cholesterol + Blood.Pressure + Smoking + Exercise.Habits,
  data   = heart_data,
  family = "binomial"
)

kable(data.frame(
  AIC                  = round(AIC(logit_model), 1),
  `Null Deviance`      = round(logit_model$null.deviance, 1),
  `Residual Deviance`  = round(logit_model$deviance, 1),
  `Degrees of Freedom` = logit_model$df.residual,
  check.names = FALSE
), row.names = FALSE, caption = "Logistic Regression Model Fit Statistics")
```

## 5.5 Odds Ratios with 95% Confidence Intervals

```{r odds-ratio-table}
lr_results <- tidy(logit_model, exponentiate = TRUE, conf.int = TRUE) %>%
  mutate(across(where(is.numeric), ~round(., 4))) %>%
  rename(
    Term         = term,
    `Odds Ratio` = estimate,
    SE           = std.error,
    `z-value`    = statistic,
    `p-value`    = p.value,
    `CI Lower`   = conf.low,
    `CI Upper`   = conf.high
  )

kable(lr_results, row.names = FALSE,
      caption = "Logistic Regression: Odds Ratios with 95% Confidence Intervals")
```

```{r odds-ratio-plot, fig.height=5}
odds_df <- tidy(logit_model, exponentiate = TRUE, conf.int = TRUE) %>%
  filter(term != "(Intercept)")

ggplot(odds_df, aes(x = reorder(term, estimate), y = estimate)) +
  geom_point(size = 2.5) +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0.25) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "red", alpha = 0.7) +
  coord_flip() +
  theme_minimal() +
  labs(
    title    = "Odds Ratios with 95% Confidence Intervals",
    subtitle = "Red dashed line indicates OR = 1 (no effect)",
    x        = "Predictor",
    y        = "Odds Ratio"
  )
```

---

# 6. Discussion

Our analysis found that none of the predictor variables were statistically significant at the
α = 0.05 level. The most notable observation was that individuals reporting any level of
`Exercise.Habits` had borderline significant odds ratios (~6.0), potentially indicating a
complex relationship with heart disease — possibly due to reverse causality or unmeasured
confounders.

Cholesterol-related variables (`Cholesterol.Level`, `Low.HDL.Cholesterol`,
`High.LDL.Cholesterol`) showed no significant association with heart disease status, which
is contrary to established clinical evidence. This may reflect limitations in the dataset's
variable encoding, sampling strategy, or population representativeness.

The logistic model had an AIC of `r round(AIC(logit_model), 0)`, suggesting modest model
fit. Future analyses could explore interaction terms, variable transformations, or
regularized regression (e.g., LASSO) to improve predictive performance and reduce
overfitting risk.

# 7. Conclusion

This study examined how cholesterol profiles and lifestyle factors relate to heart disease
risk in a dataset of 10,000 patients. While no predictor reached statistical significance
at α = 0.05, the analysis revealed patterns in exercise behavior and cholesterol markers
that warrant further investigation with larger, more clinically representative datasets.

The full pipeline — from raw data ingestion and missing value imputation, through
hypothesis testing and logistic regression, to publication-ready tables and visualizations
— is implemented in a single reproducible R Markdown document, allowing results to be
refreshed immediately whenever the underlying data are updated.
